---
title: "R Notebook"
output: html_notebook
---
## The goal of collecting this dataset

our goal is to Identify whether the chances of of a heart attack occurrence is high or low for the patient
#classification and clustering goal

Classification: We will classify target people into different groups based on appearance of muilty symptoms of Heart attack disease .This will allow us to predict the people who target to have heart attack based on certain factors .(classification done by Raghad and Taif)

Clustering: We will create a set of clusters for people who have similar symptoms. This will allow us to identify groups of people who are have the similar symptoms and target to have heart attack .(Clustring done by Sara and Leena)


We believe that by using these data mining tasks, we can build a model that can accurately predict the people target to have heart attack.

----------------------------------------------------------
modification on the pre processing done by Raghad and Taif.
----------------------------------------------------------



The dataset was sourced from the kaggle website in this URL: <https://www.kaggle.com/datasets/pritsheta/heart-attack>

## General information

1- Number of Attributes: 14 2- Number of Objects: 303 3- Type of Attributes:

| Attribute name | Data type | Description                                             | possible values                                                           |
|------------------|------------------|-------------------|------------------|
| age            | Numeric   | Age (in years)                                          | range between 29-77                                                       |
| sex            | Binary    | gender                                                  | 1 = male; 0 =female                                                       |
| cp             | Numeric   | Chest Pain type                                         | 1: typical angina,2: atypical angina,3: non-anginal pain ,4: asymptomatic |
| trestbps       | Numeric   | resting blood pressure                                  | range beteen 94-200                                                       |
| chol           | Numeric   | serum cholesterol in mg/dL                              | range between 126-564                                                     |
| fbs            | Binary    | fasting blood sugar \> 120 mg/dL (likely to be diabetic | 1=true,0=false                                                            |
| Restecg        | Binary    | Resting electrocardiogram                               | 0=normal,1=abnormal                                                       |
| Oldpeak        | Numeric   | ST depression                                           | range between 0-6.20                                                      |
| thalach        | Numeric   | heart rate achieved                                     | range between 71-202                                                      |
| exang          | Binary    | exercise induced angina                                 | 1 = yes; 0 = no                                                           |
| slope          | Numeric   | the slope of the peak exercise ST segment               | range between 0-2                                                         |
| Ca             | Numeric   | number of major vessels                                 | range between 0-4                                                         |
| Thal           | Numeric   | Thalassemia                                             | range between 0-3                                                         |
| Target         | Binary    | if the target has disease or not                        | 0 = no disease, 1 = disease                                               |

Class label:
-target of Heart Attack accurrance 
-not target of Heart Attack accurrance


-------------------
view dataset:
-------------------
```{r}
install.packages("readxl")
library(readxl)
dataset <- read_excel("C:/Users/ragha/OneDrive/سطح المكتب/New folder/Heart Attack Data Set.xlsx")
View(dataset)
```


-----------------------
Summary of dataset:
-----------------------
```{r}
nrow(dataset) 
ncol(dataset)
```

```{r}
summary(dataset$age)
summary(dataset$cp)
summary(dataset$trestbps)
summary(dataset$chol)
summary(dataset$oldpeak)
summary(dataset$thalach)
summary(dataset$slope)
summary(dataset$ca)
summary(dataset$thal)
```
The summary function provides information such as the minimum, 1st quartile, median, mean, 3rd quartile, and maximum values for each variable. It also includes the number of missing values for the thal variable.



---------------------------
Missing values and Null values
---------------------------
```{r}
is.na(dataset)
sum(is.na(dataset))
```

The output is false and 0 this indicates that the element in our dataset is not missing

--------------
Graphs
--------------
Boxplot

```{r,echo=FALSE}
install.packages("ggplot2")
library(ggplot2)

ggplot(data = dataset, aes(x = "", y = trestbps)) +
  geom_boxplot(fill = "lightgreen", color = "black") +
  labs(
    title = "Box Plot for resting blood pressure  ",
    y = "trestbps",
    x = NULL  # Remove x-axis label (since we only have one category)
  ) +
  theme_minimal()

ggplot(data = dataset, aes(x = "", y = thalach)) +
  geom_boxplot(fill = "lightpink", color = "black") +
  labs(
    title = "Box Plot for heart rate achieved ",
    y = "heart rate",
    x = NULL  # Remove x-axis label (since we only have one category)
  ) +
  theme_minimal()

ggplot(data = dataset, aes(x = "", y = oldpeak)) +
  geom_boxplot(fill = "purple", color = "black") +
  labs(
    title = "Box Plot for ST depression",
    y = " ST depressione",
    x = NULL  # Remove x-axis label (since we only have one category)
  ) +
  theme_minimal()
ggplot(data = dataset, aes(x = "", y =chol)) +
  geom_boxplot(fill = "lightyellow", color = "black") +
  labs(
    title = "Box Plot for cholestero",
    y = "chol",
    x = NULL  # Remove x-axis label (since we only have one category)
  ) +
  theme_minimal()

```

we can also detect summary of each attribute from the box plot , and we use box plot to detect  the outlier.

Histogram
```{r}
hist(dataset$chol)
```

The graph represents the frequency of cholesterol in the data set. After observing the values, we concluded that most of the values fall in the range higher than the normal range, as the normal range falls between  125-200.

Scatter plot
```{r}
library(ggplot2)
gg <- ggplot(dataset, aes(x=chol, y=trestbps)) + geom_point(aes(col=target, size=oldpeak)) + labs(subtitle="trestbps Vs chol", y="trestbps", x="chol", title="Scatterplot", 
       caption = "Source: midwest", bins = 30)
plot(gg)
```
The scatter plot represents a comparison between the variables 'trestbps' and 'chol'. The plot displays each data point as a dot on the chart, where the x-axis represents 'trestbps' values and the y-axis represents 'chol' values.

In this scatter plot, we can observe a pattern or trend, indicating that there might be a correlation between 'trestbps' and 'chol'. This could potentially mean that changes in one variable might be associated with changes in the other variable.

The range of 'trestbps' values is from 100 to 175, while the range of 'chol' values is from 230 to 320. This indicates that there are both positive and negative relationships between these two variables.


------------
Bar plot
------------
 There is no point in using bar plots because the frequency of one symptom (ex: high blood pressure alone ) is not enough to decide whether the patients are likely to have a heart attack or not

--------------
 outliers:
--------------

Detecting the outliers:
```{r}
  boxplot.stats (dataset$chol)$out
  boxplot.stats (dataset$trestbps)$out
  boxplot.stats (dataset$oldpeak)$out
  boxplot.stats (dataset$thalach)$out
```


Removing the outliers:
```{r}
outliers <- boxplot (dataset$chol, plot=FALSE) $out
 dataset <- dataset [-which(dataset$chol %in% outliers),]
 boxplot.stats (dataset$chol) $out
```

The outlier in the attribute wasn’t a real Outlier and if we remove all the Outliers, our dataset affected and become small so we remove the Outlier from the attribute control that we noticed it is real Outlier.

--------------------------------
Transformation and Normalization
--------------------------------
Discretization:

```{r,echo=FALSE}
dataset$age<-cut(dataset$age, breaks = seq(29, 77, by = 10),right=TRUE)
```
In this process, we divide the age values into five intervals with equal width. This allows us to simplify and classify the data, making it more suitable for later use in our model. By discretizing the age values, we effectively group them into five meaningful categories. This transformation will result in a clearer and more organized data set that can be easily understood and utilized in subsequent stages of our model.



Normalization:
```{r}
normalize <- function(x) {
  return((x - min(x)) / (max(x) - min(x)))
}

dataset$trestbps <- normalize(dataset$trestbps)
dataset$chol <- normalize(dataset$chol)
dataset$thalach <- normalize(dataset$thalach)
dataset$oldpeak <- normalize(dataset$oldpeak)
dataset$cp <- normalize(dataset$cp)
dataset$oldpeak <- normalize(dataset$oldpeak)
dataset$slope <- normalize(dataset$slope)
dataset$thal <- normalize(dataset$thal)
```

In order to ensure comparability across individuals and eliminate biases resulting from differences in measurement units or scales, we have decided to normalize the following attributes: trestbps, chol, thalach, oldpeak, cp, oldpeak, slope,and thal.


------------------------------------
feature selection
------------------------------------
we did not applly feature selection beacause the number of attribute is small and we will  neeed them
------------------------------------

##Correlation Analysis: We will find the correlation between each attributes and the class label (target). For the nominal data we will use chi-square and for the numeric data we will use correlation coefficient. This will help us to determine the most important and correlated attributes to the target.
```{r}
####Sex:

csex=chisq.test(dataset$sex , dataset$target)
print(csex)
####Chest pain type (cp):

ccp=chisq.test(dataset$cp , dataset$target)
print(ccp)
####Fasting blood sugar (fbs):

cfbs=chisq.test(dataset$fbs , dataset$target)
print(cfbs)
####Resting electrocardiographic result (restecg):

crestecg=chisq.test(dataset$restecg , dataset$target)
print(crestecg)
####Exercise induced anginal (exang):

cexang=chisq.test(dataset$exang , dataset$target)
print(cexang)
####The slope of the peak exercise ST segment (slope):

cslope=chisq.test(dataset$slope , dataset$target)
print(cslope)
####Number of major vessels (ca):

cca=chisq.test(dataset$ca , dataset$target)
print(cca)
####A blood disorder (thal):

cthal=chisq.test(dataset$thal , dataset$target)
print(cthal)
####Age: The age attribute after discretization.

cage= chisq.test(dataset$age, dataset$target)
print(cage)
####Resting blood pressure (trestbps)

ctrestbps=chisq.test(dataset$trestbps ,dataset$target)
print(ctrestbps)

```
|  Attribute name | Chi-square value|Degree of freedom|Alpa|             
  |-----------------|-----------------|-----------------|-----------------|
  | A blood disorder (thal) | 85.304|3|2.2e-16|
   | Chest pain type (cp)    | 81.686 |3| 2.2e-16| 
  | Number of major vessels (ca)| 74.367|4|2.712e-15|
  | Exercise induced anginal (exang)| 55.945 |1|7.454e-14|
  |The slope of the peak exercise ST segment (slope)| 47.507|2|4.831e-11|
  | Sex            | 22.717       | 1| 1.877e-06|
  |Resting blood pressure (trestbps)|  47.706|48|0.4848|
  | Resting electrocardiographic result (restecg)| 10.023 |2|0.006661|     
  |Age(after discretization)| 15.689  |3|0.001313|      
  | Fasting blood sugar (fbs)    | 0.10627   |1|0.7444|

    
###Chi-square Results: We will sort the Chi-square values from the highest to the lowest:

1-A blood disorder (thal) with value 85.304.
2-Chest pain type (cp) with value 81.686.
3-Number of major vessels (ca) with value 74.367 .
4-Exercise induced anginal (exang) with value 55.945 .
5-resting blood pressure (trestbps) with value 47.706 .
6-The slope of the peak exercise ST segment (slope) with value 47.507 .
7-Sex with value 22.717 .
8-Age with value 15.689 .
9-Resting electrocardiographic result (restecg) with value 10.023.
10-Fasting blood sugar(fbs) with value 0.10627 .
All of the attributes is dependent to the class lable (target) except Fasting blood sugar(fbs) since the p-value is higher than chi-square value (chi-square = 0.10627 < 0.7444=p-value).   
 

----------------------
###Correlation coefficient for numeric data:
----------------------
```{r}
####Serum cholestoral (chol):

cchol=cor(dataset$chol ,dataset$target)
print(cchol)
####Maximum heart rate achieved (thalach):

cthalach=cor(dataset$thalach ,dataset$target)
print(cthalach)
####ST depression indicated by exercise relative to rest (oldpeak):

coldpeak=cor(dataset$oldpeak ,dataset$target)
print(coldpeak)
```
###Correlation coefficient Results: The highest correlation coefficient value is ST depression indicated by exercise relative to rest (oldpeak) which is negatively correlated (-0.430696) with the target. Then, the value of maximum heart rate achieved (thalach) which is positively correlated (0.4217409) with the target. Lastly, the value of Serum cholestoral (chol) which is negatively correlated (-0.08523911) with the target.

From the previous results we decided to delete the Fasting blood sugar (fbs) since it is independent from the class label (target) which means that fasting blood sugar (fbs) doesn't affect our class label (target).

```{r}
dataset <- dataset[, -which(names(dataset) == "fbs")]
head(dataset)
```


```{r}
ncol(dataset)
nrow(dataset)
```
After printing a sample from our new data set we notice that the attribute decrease from 14 to 13 since we remove Fasting blood sugar (fbs) and the rows remain the same.


print the final preprocessed dataset
------------------------------------
```{r}
print(data)
```


